### 2017-12-08 paper/sexSelectionSpacing-ver1.tex
Reorganization and shortening of Estimation Strategy
First present probabilities of boy and girl at each time period.
Then discuss the problem of interpretation of variables, followed
by survival curve and how to address differences in parity
progression likelihood.
Finally, show how to calculate percentage boys and discuss the
graphs associated with those.

The main downside of this set-up is that direct interpretation of the estimated 
coefficients is challenging;
the coefficients show the change in hazards relative to the base outcome, 
here no birth, rather than simply the hazard of an event.
Hence, a positive coefficient does not necessarily imply that the associated 
exit state becomes likely as the explanatory variable increases because the 
probability of the other exit state may increase even more \citep{thomas96}.
It is, however, straightforward to calculate the predicted probabilities of 
having a boy and of having a girl for each $t$ within a spell, conditional on 
a set of explanatory variables and not having had a child before that period.
The predicted probability of having a boy in period $t$ for a given set of 
explanatory variable values, $\mathbf{Z}_k$ and $\mathbf{X}_k$, is
\begin{equation}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})},
\label{eq:probability_boy}
\end{equation}
and the predicted probability of having a girl is
\begin{equation}
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}.
\label{eq:probability_girl}
\end{equation}

[survival curve here]

With these two probabilities is it easy to calculate, for each $t$, the estimated
percentage of children born that are boys, $\hat{Y}$, 
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100,
\label{eq:probability_son}
\end{equation}
together with the associated confidence interval for given values of explanatory 
variables.%
\footnote{
Within each period $1-P(b_{t})-P(g_{t})$ is the probability of not having a birth in 
period $t$.
}

For ease of exposition the procedure is presented here in two steps, but the actual
calculation of the percent boys is done in one step with the 95 percent confidence 
interval calculated using the Delta method.
Results are presented as the estimated percent boys born by length of birth spacing
using graphs.%
\footnote{
The parameter estimates are available on request.
}
For each graph, the extent to which the percent boys is statistically 
significantly above the natural sex ratio indicates the use of sex selection.

[TK  may need to combine with and without sons to show differences in spacing 
over time; maybe it should also have predicted average/median spacing with some 
confidence intervals to test]

The distribution of spacing is captured by the survival curve, which shows 
the probability of not having had a birth yet by spell duration.
The survival curve at time $t$ is 
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t 
\left( 
	1- \left(P(b_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) 
	+ P(g_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) \right) 
\right),
 \label{eq:survival}
\end{equation}
or equivalently
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t
\left(
\frac{ 1 }
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{ld}'\mathbf{Z}_{kd} + \beta_l'\mathbf{X}_{k})}
\right).
\end{equation}

[The problem with comparing survival curves]

In principle, in order to understand how birth spacing changes in response
to the introduction of prenatal sex determination one could simply compare survival 
curves between different sex compositions of prior births over time.
If the idea that the availability of sex-selective abortion increases the
spacing more after the birth of girls than after the birth of boys is 
correct, we should see the two survival curves become closer to each other
and we might even be able to directly observe that the survival curve for prior
child being a girl is above the survival curve for prior child being a boy,
instead of substantially below as was generally the case before prenatal
sex determination become widely available.

There are, however, two (interrelated) problems.
First, the probability that a woman will progress to a next birth  
is not the same across different sex compositions of prior children.
Specifically, one of the ways that past research has tried to establish
son preference in fertility behavior is to compare parity progression
rates for different sex compositions [TK references - see footnote 1].
The idea was that if there is son preference a couple would be less likely
to progress to a next birth if they have one or more son already for a
given number of prior births.

The fact that the probability of having a next birth also depends on the
sex composition complicates the comparison of survival curves. 
A lower parity progression probability will naturally make the survival
curve flatter.
This means that the survival curve for a girl composition will be more
likely to be below the boy composition.
Even if the use of sex selection is substantial enough to move the
survival curve for girl composition to the right of the boy
composition the lower parity progression probability for boy 
composition means that the survival curves would eventually have to
cross.
This makes it difficult---if not impossible---to design a direct test
of how the survival curves compare.

The problem of different parity progression rates across different
sex composition is exacerbated by falling fertility over time.
This is because there are now more parities where they may be a 
substantial differences in parity progression rates and because
the difference in parity progression for a given parity will 
become larger if there is son preference.
As the parity progression probability decreases it becomes impossible
to capture the time until a certain survival point (like 0.5 of women
having had a child; the survival curve may never actually reach this
point).

In addition, it is possible that falling fertility also generically 
leads to longer spacing between births.
This means that we have to separate out which part of the longer
spacing comes from falling fertility (and therefore should also
affect boy composition) and which parts comes from increased use
of sex selection.

One solution to this problem is to condition on the parity progression
rates.
That is, instead of comparing say how long it takes for 50\% of women
who began the spell to have the next birth across boy and girl composition, 
which would work well if the next birth was nearly universal, we 
could calculate how long it takes for 50\% of those women who will 
eventually have a next child to have that child based on the estimated
survival curves.
This number can then be compared across sex compositions to 
understand if the spell length for girl composition increases
more than for boy composition with the introduction of sex selection.
Obviously, this should be calculated for different percentages
and the easiest way to interpret the result is to graph survival
conditional on parity progression occurring.
All compositions therefore begin at 100\% and end at 0\% because
we condition on having the next child.

Note, this is not the same as simply calculating the average 
\emph{observed} spell lengths among women who have a given parity
child in the survey because that number does not take into account
the censoring of some spells that will eventually lead to a birth.

@@

>>>>

The main downside of this set-up is that direct interpretation of the estimated 
coefficients is challenging;
the coefficients show the change in hazards relative to the base outcome, 
here no birth, rather than simply the hazard of an event.
Hence, a positive coefficient does not necessarily imply that the associated 
exit state becomes likely as the explanatory variable increases because the 
probability of the other exit state may increase even more \citep{thomas96}.
It is, however, straightforward to calculate the predicted probabilities of 
having a boy and of having a girl for each $t$ within a spell, conditional on 
a set of explanatory variables and not having had a child before that period.
The predicted probability of having a boy in period $t$ for a given set of 
explanatory variable values, $\mathbf{Z}_k$ and $\mathbf{X}_k$, is
\begin{equation}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})},
\label{eq:probability_boy}
\end{equation}
and the predicted probability of having a girl is
\begin{equation}
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}.
\label{eq:probability_girl}
\end{equation}
Within each period the probability of not having a birth in period $t$
is $1-P(b_{t})-P(g_{t})$.

The distribution of spacing is captured by the survival curve, which shows 
the probability of not having had a birth yet by spell duration, for
a given set of explanatory variables.
The survival curve at time $t$ is 
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t 
\left( 
	1- \left(P(b_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) 
	+ P(g_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) \right) 
\right),
 \label{eq:survival}
\end{equation}
or equivalently
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t
\left(
\frac{ 1 }
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{ld}'\mathbf{Z}_{kd} + \beta_l'\mathbf{X}_{k})}
\right).
\end{equation}

An important issue is that the probability of ever having a next birth varies 
across groups.
For example, as the parity progression literature shows, couples with stronger 
son preference are less likely, for a given number of prior births, to ever have 
a next birth if they already have one or more sons.
Direct comparison of standard survival curves therefore tells us little 
about how the spread of sex selection affect birth spacing across, for
example, different sex compositions of prior children.
To overcome this problem I condition on parity progression when examining
birth spacing using the survival curves.
This allows me to calculate measures, such as median spell duration, for
the women with a given set of characteristics who will eventually give
birth in a spell.%
\footnote{
It is important to note that this is not the same as simply calculating 
the median \emph{observed} spell lengths among women who have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
}
This number can then be compared across groups to understand how birth
spacing changes with the introduction of sex selection.
Obviously, this approach only works if the spell length covered is 
sufficiently long that only very few women would be likely to give
birth later.
In addition to calculating conditional spell length for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

Finally, with the predicted probabilities of having a boy and of having 
a girl for each $t$ within a spell, is it easy to calculate the estimated 
percentage of children born that are boys, $\hat{Y}$, at each $t$  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting across all $t$ 
gives the predicted percent boys born over the entire spell.
In addition to the predicted percent boys over the entire spell, 
I also present graphs of how the percentage boys born vary across time
within a spell, together with the associated confidence interval for given 
values of explanatory variables calculated using the Delta method.
These are combined with standard survival curves which shows the ``weight''
that each period should be given, or, in other words, how many births take
place in each period.

@@

Results are presented as the estimated percent boys born by length of birth spacing
using graphs.
For each graph, the extent to which the percent boys is statistically 
significantly above the natural sex ratio indicates the use of sex selection.

[TK not sure where this fits in]
In addition, it is possible that falling fertility also generically 
leads to longer spacing between births.
This means that we have to separate out which part of the longer
spacing comes from falling fertility (and therefore should also
affect boy composition) and which parts comes from increased use
of sex selection.
